Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-04-30 text/back_end_KS.qmd
readded deleted LA rq2 wtf

The enduring segregated structure of the Chicago metropolitan area from 1980 to 2020 analyzed above plays a significant role in how Geomarkets shape the racial/ethnic and socioeconomic composition of prospect pools. Reflecting the larger metropolitan area dynamics, more central Geomarkets such as the City of Chicago and South and Southwest Suburbs consistently contribute disproportionately larger shares of Black, Hispanic, and first-generation college prospects, especially at lower SAT score thresholds. On the other hand, Geomarkets in the suburban ring of the metropolitan area like the Chain of Lakes, North Shore, and Evanston and Skokie contribute larger proportions of White and Asian prospects, particularly as SAT scores increase. However, the City of Chicago stands out in playing crucial role in fostering college access for first-generation Asian college students, contributing such prospects to the overall pool at significantly higher, disproportionate rates than any other Geomarket.

@@

The Dallas-Fort Worth metroplex experienced substantial demographic growth since Geomarket boundaries were dilineated by @RN4982 in the early 1980s through 2020. The metropolitan area's unique urban structure of a dual central cities surrounded by concentric suburbs has shaped patterns of population growth across racial/ethnic, economic, and spatial shifts in the Geomarket lanscape. While all six Geomarkets experienced population increases, the most significant growth occurred in the outer suburban areas, reflecting ongoing urban sprawl. However, all Geomakets have experienced substantial growth in Hispanic, Black, and Asian populations leading to only two of the six Dallas-Fort Worth Geomarkets remaining predominantly White (Collin and Rockwall Counties and West of Dallas/Fort Worth Metroplex. On the other hand, the metropolitan area has experienced widening disparities in income, poverty, and education attainment across Geomarkets, especially between the historically affluent Collin and Rockwall Counties and the less economically advantaged inner-core Geomarkets (City of Fort Worth and City of Dallas). Together, these trends underscore the growingly racial/ethnic diverse but widening economic inequality within the evolving Geomarket landscape of the Dallas-Fort Worth metroplex.

@@

@@

### Geomarket Changes Over Time \[Need better RQ1 heading\]

@@

@@

>>>>

The enduring segregated structure of the Chicago metropolitan area from 1980 to 2020 analyzed above plays a significant role in how Geomarkets shape the racial/ethnic and socioeconomic composition of prospect pools. Reflecting the larger metropolitan area dynamics, central Geomarkets that are predominantly comprised of low-income communities of color such as the City of Chicago and South and Southwest Suburbs consistently contribute disproportionately larger shares of Black, Hispanic, and first-generation college prospects, especially at lower SAT score thresholds. On the other hand, predominantly White and affluent Geomarkets in the suburban ring of the metropolitan area like the Chain of Lakes, North Shore, and Evanston and Skokie contribute larger proportions of White and Asian prospects, as well as non-first generation college students. However, the City of Chicago stands out in playing crucial role in fostering college access for first-generation Asian college students, contributing such prospects to the overall pool at significantly higher, disproportionate rates than any other Geomarket.

@@

The Dallas-Fort Worth metroplex experienced substantial demographic growth since Geomarket boundaries were dilineated by @RN4982 in the early 1980s. While all six Geomarkets experienced population increases, the most significant growth occurred in the outer suburban areas, reflecting ongoing urban sprawl. Such growth is largely attributed to increases in Hispanic, Black, and Asian populations leading to only two of the six Dallas-Fort Worth Geomarkets remaining predominantly White (Collin and Rockwall Counties and West of Dallas/Fort Worth Metroplex. On the other hand, the metropolitan area has experienced widening disparities in income, poverty, and education attainment across Geomarkets, especially between the historically affluent Collin and Rockwall Counties and the less economically advantaged inner-core Geomarkets (City of Fort Worth and City of Dallas). Together, these trends underscore the growingly racial/ethnic diverse but widening economic inequality within the evolving Geomarket landscape of the Dallas-Fort Worth metroplex.

### Composition of Students Lists by Geomarkets

@@

Analyses of student list orders across the Dallas-Fort Worth metropolitan area reveals patterns of racial/ethnic and socioeconomic stratification in Geomarket contributions. As Dallas-Fort Worth became increasingly diverse from 1980 to 2020, a more balanced racial/ethnic distribution emerged across the metropolitan area's Geomarket landscape. However, these patterns did not translate to racial/ethnic composition of student prospect pools. Suburban Geomarkets like Collin and Rockwall Counties and the West of Dallas/Fort Worth Metroplex yielded a disproportionately larger share of Asian prospects, whereas central Geomarkets such as the City of Dallas, Dallas County, and Irving, Arlington, and Grand Prairie contributed a disproportionately larger share of Hispanic and Black prospects. These racial/ethnic disparities become significantly more pronounced, particularly in the disproportionate share of Asian prospects, as PSAT score thresholds increase (see Appendix @fig-dallas-rq2-lowSAT-race and @fig-dallas-rq2-highSAT-race). Similar Geomarket contribution patterns are also evident by first-generation college student status, with Collin and Rockwall Counties and the West of Dallas/Fort Worth Metroplex providing larger shares of non-first generation college student prospects, and central Geomarkets (City of Dallas, City of Fort Worth, Dallas County, and Irving, Arlington, and Grand Prairie) providing disproportionately larger shares of first-generation college student prospects. Moreover, central Geomarkets are particularly important for including Black (City of Dallas and City of Fort Worth), Hispanic (City of Dallas, City of Fort Worth, Dallas County, and Irving, Arlington, and Grand Prairie), and Asian (City of Fort Worth and Irving, Arlington, and Grand Prairie) first-generation college students. While Collin and Rockwall Counties and the West of Dallas/Fort Worth Metroplex Geomarkets still contribute a notable overall number of Black, Hispanic, and Asian prospects, such students are predominantly not first-generation college students.

@@

### Decentralized Diversity and Income in a Multi-Nucleo in the Los Angeles Geomarket Landscape

@@

The Los Angeles metropolitan area exemplifies a complex multi-nuclei urban model, characterized by decentralized population growth and increasing demographic diversity across its 11 Geomarkets. From 1980 to 2020, the region’s population expanded significantly leading to overall declines in the White population and increases in Hispanic and Asian populations across Geomarkets. West Los Angeles and West Beach is the only majority White Geomarket in 2020. Socioeconomic indicators reveal enduring disparities, especially in historically lower-income Geomarkets like South and South Central Los Angeles, East Los Angeles, and Hollywood and Wilshire. Meanwhile, Geomarkets such as West Los Angeles and West Beach became increasingly affluent and highly educated, highlighting growing socioeconomic and educational divides. Given these demographic shifts within the context of a multi-nuclei urban model, the Los Angeles metropolitan area presents a distinctive Geomarket landscape case in which racial/ethnic enclaves across a range of various socioeconomic levels are distributed across multiple Geomarkets rather than concentrated in one or two.

@@

@fig-la-rq2-midSAT-race presents the race/ethnicity of all prospects targeted in the middle PSAT score range across Geomarket contributions for the Los Angeles metropolitan area. For the more than 2,500 White prospects in this pool, we see the San Fernando Valley - West (28% versus 20% for all prospects) and the West Los Angeles and West Beach (12% versus 7% for all prospects)Geomarkets contributing disproportionately larger shares. Larger disparities across Geomarket contributions were evident for Asian, Black, and Hispanic students. The Glendale and Pasadena (17% versus 14% for all prospects), East Los Angeles (11% versus 5% for all prospects) and Covina and West Covina (17% versus 11% for all prospects) Geomarkets contributed a disproportionate larger share of Asian students, whereas nearly all other Geomarkets contributed disproportionately smaller or nearly equal shares.

Representational patterns across Geomarkets are reversed for Black and Hispanic prospects in @fig-la-rq2-midSAT-race. South and South Central Los Angeles and Riverside, San Bernadino, and Ontario Geomarkets contributed larger proportions of Black and Hispanic prospects relative to their contributions to the overall prospect pool in the Los Angeles metropolitan area. On the other hand, the San Fernando Valley- West and the Glendale and Pasadena Geomarket contributed smaller proportions of Black and Hispanic Prospects. However, the West Los Angeles and West Beach (10% versus 7% for all prospects), Hollywood Wilshire (7% versus 5% for all prospects), and the Long Beach (8% versus 5% for all prospects) Geomarkets contributed a larger share of Black prospects, whereas Covina and West Covina contributed a larger share of Hispanic students. All other Geomarkets contributed nearly equal shares of Black and Hispanic prospects.

Contributions of each Geomarket to the pool of prospects from the Los Angeles metropolitan area based on first-generation college student status are presented in @fig-la-rq2-midSAT-ses. Only 889 of the more than 8,000 total prospects (11%) in this middle PSAT score range (1190-1260) pool were first-generation college students whose parents did not attend college. Only three of the 11 Geomarkets contributed substantially larger shares of these first-generation prospects: East Los Angeles (35%), South and South Central Los Angeles (47%), and Riverside, San Bernardino, and Ontario (15%). However, disproportionate contributions become more spread across Geomarkets for the nearly 1,436 of the more than 8,000 total prospects (18%) who were first-generation college students whose parents attended but did not complete their degree. Overall, six of the 11 Geomarkets contributed larger shares of these prospects: Hollywood & Wilshire (25%), East Los Angeles (26%), South and South Central Los Angeles (28%), Long Beach (25%), Covina and West Covina (22%), and Riverside, San Bernardino, and Ontario (22%). This leads to the remaining Geomarkets - San Fernando Valley- West (80%), San Fernando Valley- East (78%), Glendale and Pasadena (77%), West Los Angeles & West Beach (87%), South Bay (81%) - contributing disproportionately larger shares of not first-generation college students relative to the overall prospect pool (71%).

Lastly, @fig-la-rq2-midSAT-combo presents Geomarket contributions to prospect pools by both first-generation college student status and race/ethnicity in the Los Angeles metropolitan area. Among all White prospects (n=2,511) whose contact information was purchased in this middle PSAT score range, approximately 3% were first-generation college students whose parents did not attend college, 11% were first-generation college students whose parents attended some college but did not complete their degree, and 86% were not first-generation college students. @fig-la-rq2-midSAT-combo shows all Geomarkets contribute a nearly proportionate share of prospects by first-generation college students (across both no college and some college parents) relative to the overall pool. The South and South Central Los Angeles Geomarket stands out as one exception contributing a relatively larger share of White first-generation college students whose parents did not attend college (17%) and whose parent attended some college but did not complete their degree (33%). Other Geomarket that contributed signifcantly larger shares of White first-generation prospects whose parents did not complete their degree include Long Beach (17%), and Riverside, San Bernardino, and Ontario (19%).

Geomarkets contributions to first-generation college students was mixed across other racial/ethnic categories in @fig-la-rq2-midSAT-combo. For Asian prospects (11% first-generation no college, 21% first-generation some college, and 68% not first-generation), the Hollywood & Wilshire (12%) and East Los Angeles (36%) Geomarkets contributed larger proportions of first-generation college students whose parents did not attend college. The San Fernando Valley- East (28%), Hollywood & Wilshire (35%), East Los Angeles (25%), South and South Central Los Angeles (44%), Long Beach (27%), and Covina and West Covina (25%) Geomarkets also contributed larger proportions of first-generation college students whose parents attended college but did not complete their their degree.

For Black prospects (5% first-generation no college, 20% first-generation some college, and 74% not first-generation), Geomarkets that disproportionately larger proportions of first-generation college students whose parents did not attend college include South Bay (11%), South and South Central Los Angeles (21%), Long Beach (7%), and Covina and West Covina (9%). The West Los Angeles and West Beach (26%), Hollywood & Wilshire (36%), East Los Angeles (33%), Long Beach (40%), and Covina and West Covina (27%) Geomarkets also contributed larger proportions of Black first-generation college students whose parents attended college but did not complete their their degree.

Finally, Hispanic prospects in this middle range PSAT order were on average 25% first-generation college students whose parents did not attend college, 25% first-generation college students whose parents attended some college but did not complete their degrees, and 50% were not first-generation college students. The Hollywood & Wilshire (30%), East Los Angeles (39%), South and South Central Los Angeles (54%), and Riverside, San Bernardino, and Ontario (31%) contributed disproportionate shares of Hispanic first-generation college students whose parents did not attend college. The East Los Angeles (33%), South and South Central Los Angeles (28%), Long Beach (30%), and Riverside, San Bernardino, and Ontario (27%) also contributed larger Hispanic first-generation college students whose parents attended college but did not complete their their degree.
